Understanding Factors Contributing to Application Approval in Urban Planning Decisions (2016-2024)

### 2024-12-03 paper/paper.qmd
Revise paper

---
title: "Understanding Factors Contributing to Application Approval in Urban Planning Decisions (2016-2024)"
subtitle: "Analysis using data from Toronto’s Committee of Adjustment applications using Bayesian Logistic Regression"
author: Angel Xu
thanks: "Code and data are available at: [https://github.com/Anjojoo/Committee-of-Adjustment-Applications](https://github.com/Anjojoo/Committee-of-Adjustment-Applications)."
date: today
date-format: long
abstract: "This paper analyzes urban planning application approvals submitted to Toronto's Committee of Adjustment from 2016 to 2024, focusing on key predictors such as application type, submission year, and planning district using Bayesian logistic regression. The analysis shows that consents are more likely to be approved than minor variances, with approval rates generally increasing over time, particularly from 2017 onward, though fluctuations are observed in certain years. Geographic disparities were identified, with North York showing the highest likelihood of approval, while Scarborough faced stricter reviews. These findings enhance understanding of urban planning decision-making processes, offering insights into temporal trends, geographic variability, and areas for equitable policy adjustments."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

This study aims to estimate the factors influencing the approval decisions for urban planning applications in Toronto. By analyzing variables such as the type of application (minor variance or consent), the year of submission, and the planning district, this study seeks to understand how these elements shape the decision-making process. The primary objective is to determine the causal effects of these factors on approval outcomes, thereby offering insights into the dynamics of urban planning and regulatory processes.

@@

The Committee of Adjustment plays a key role in Toronto’s urban planning ecosystem. It ensures that deviations from zoning bylaws align with the city’s broader development objectives, maintaining a balance between flexibility for property owners and adherence to long-term planning goals. The decisions made by the committee are influenced by various factors, including the type of application, geographic considerations (e.g. planning districts), and the timing of submission. Understanding these decisions provides insight into urban planning dynamics and the regulatory challenges faced by a growing city like Toronto.

@@

The dataset thus provides a structured lens through which to examine urban planning decisions. By distilling real-world phenomena—such as zoning adjustments and land use approvals—into discrete variables, the dataset enables systematic analysis of trends and relationships within Toronto’s regulatory framework. This structured approach ensures that the dataset is both practical for analysis and representative of the nuanced processes underlying urban development.

@@

#### Plot the distribution of `decision` ####
ggplot(analysis_data, aes(x = factor(decision))) +
  geom_bar(fill = "lightblue", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(
    x = "Decision (0 = Refused, 1 = Approved)",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

@@

#### Plot the distribution of `application_type` ####
ggplot(analysis_data, aes(x = factor(application_type))) +
  geom_bar(fill = "lightgreen", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(
    x = "Application Type",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

@@

The year variable represents the year of submission for each application in the dataset. It provides insight into temporal trends in applications submitted to the Committee of Adjustment. This variable is instrumental in identifying changes in the volume of applications over time, which may reflect policy shifts, urban development trends, or external factors like economic conditions.

@@

#### Plot the distribution of `year` ####
ggplot(analysis_data, aes(x = factor(year))) +
  geom_bar(fill = "skyblue", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(
    x = "Year",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

@@

The year variable is likely related to the decision outcome and application type. As approval rates may have varied over time due to changing city planning policies or committee priorities. Besides, temporal trends may reveal shifts in the proportion of Minor Variances (MV) versus Consents (CO), reflecting evolving development demands.

@@

The planning district variable identifies the geographic district within Toronto where an application was submitted. The city is divided into four planning districts, each with its own development characteristics and zoning regulations. These districts are crucial for understanding spatial trends in application submissions and outcomes, as local policies and priorities can influence decision-making processes.

@@

#### Plot the distribution of `planning_district` ####
ggplot(analysis_data, aes(x = factor(planning_district))) +
  geom_bar(fill = "coral", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  labs(
    x = "Planning District",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

@@

After observing the data, the posterior distributions of $\beta$ are obtained by combining these priors with the likelihood of the observed outcomes. The resulting posterior distributions provide updated beliefs about the coefficients, incorporating both prior knowledge and evidence from the data. This approach allows for robust inference, offering a clear quantification of parameter uncertainty and supporting more nuanced interpretations of the predictors' effects.

@@

The Bayesian logistic regression model is an appropriate choice for analyzing how application type, submission date, and planning district influence the likelihood of application approval. This choice is driven by the binary nature of the response variable, the model's ability to incorporate prior knowledge, and its probabilistic framework, which quantifies uncertainty more effectively than frequentist alternatives. Logistic regression naturally constrains predicted probabilities to the [0,1] interval, ensuring interpretability as likelihoods of approval, while the Bayesian approach enhances flexibility and provides richer insights through posterior distributions.

@@

This Bayesian model relies on several assumptions that ensure the validity of its predictions and the transparency of its coefficients. While these assumptions are generally less restrictive than those for simpler models like linear regression, they remain critical to the robustness of the results. The key assumptions are listed below:

@@

The result highlights how the predictors influence the likelihood of application approval while controlling for other variables. The coefficients, presented as log-odds, provide insights into both the direction and magnitude of each predictor’s impact on the approval decision.

@@

The year variables reveal significant temporal variations in approval likelihood, using 2016 as the reference year. These coefficients highlight how approval dynamics have shifted over time. For instance, in 2017, the coefficient of 0.74 indicates a 109% increase in the odds of approval compared to 2016 ($e^{0.74} \approx 2.09$), which suggests a substantial policy or procedural shift. In 2018, the strongest positive effect is observed, with a coefficient of 0.96, reflecting a 161% increase in odds relative to 2016 ($e^{0.96} \approx 2.61$). However, in 2019, the coefficient drops to 0.22, representing a smaller increase of 25% in odds ($e^{0.22} \approx 1.25$), suggesting that this period may have faced stricter reviews or procedural adjustments. By 2020, the coefficient rebounds to 0.56, translating to a 75% increase in odds ($e^{0.56} \approx 1.75$).

@@

In summary, the model reveals key patterns in how application type and submission year influence the likelihood of approval. Minor variance applications face slightly lower odds of approval compared to consents, while temporal trends indicate an overall increase in approval rates from 2016 to 2020. These findings provide a foundation for understanding the dynamics of approval decisions and highlight areas for further exploration, such as geographic variability or interactions between predictors.

@@

This paper investigates the factors influencing application approvals submitted to the City of Toronto's Committee of Adjustment, using a Bayesian logistic regression model. By analyzing predictors such as application type, submission year, and planning district, the study provides insights into the dynamics of urban planning decisions. The model captures temporal trends, spatial variability, and procedural differences, offering a structured understanding of approval outcomes. Visualizations, including coefficient plots and effect plots, complement the analysis, helping to interpret the relationships between predictors and approval likelihood.

@@

One key finding is the temporal trend in approval rates. The analysis reveals that approval odds have increased steadily over time, peaking in 2018. This trend may reflect evolving planning policies, greater efficiency in committee processes, or a shift in zoning priorities. The data suggests that the Committee of Adjustment has become more permissive in granting approvals, which could indicate a growing alignment between applications and city planning objectives. This result highlights how administrative processes adapt over time to urban development needs.

@@

Spatial variability plays a significant role in application approvals. The findings show that North York consistently exhibits the highest approval rates, likely due to its prominence as a development hub with infrastructure supporting large-scale projects. Conversely, Scarborough has the lowest odds of approval, suggesting stricter zoning enforcement or less alignment with development objectives in this area. This geographic disparity underscores the influence of local contexts and priorities on urban planning decisions, revealing how different districts balance growth and zoning compliance.

@@

While this paper provides insights into the factors influencing application approvals, there exists several limitations. These weaknesses cover data-related issues, methodological constraints, and broader contextual considerations.

1. **Data Limitations**
The analysis relies solely on the dataset provided by the City of Toronto's Committee of Adjustment, which, while complete in its coverage of applications, lacks certain critical variables. Key contextual factors, such as community objections, project complexity, or the presence of political or economic pressures, are not captured. For instance, the dataset does not provide information about why certain applications are refused or approved beyond the recorded decision, limiting the ability to analyze qualitative or situational influences.

@@

3. **Simplified Representation of Approval Dynamics**
The approval process involves a combination of bureaucratic procedures, stakeholder negotiations, and technical considerations that are not fully captured in this analysis. For example, public hearings, lobbying efforts, and input from local communities likely play a significant role in shaping decisions but are absent from the dataset. This simplification means that the model captures only part of the story, potentially overlooking nuanced dynamics in urban planning and governance.

@@

One critical next step is to expand the dataset by incorporating additional variables that capture contextual and qualitative factors influencing decisions. Data on community feedback and objections during public hearings is a good choice since it could provide insights into how local resistance or support impacts approval likelihood. Information on project complexity, such as the scale, type, and potential impact of proposals, could reveal patterns in decision-making. Besides, access to committee meeting notes or further exploration in meeting minutes would help clarify the reasoning behind specific approvals or refusals. Furthermore, integrating economic and political context, such as local economic indicators or changes in political leadership, would offer a broader understanding of approval trends.

The current model assumes additive relationships between predictors and approval likelihood, which may oversimplify real-world dynamics. Future research could explore nonlinear relationships, such as whether certain predictors have diminishing or exponential effects on approval rates. Interaction effects between predictors, such as the interplay between planning districts and application types, could also provide more nuanced insights. For instance, minor variances may be treated differently in densely populated urban areas compared to suburban districts.

@@

Despite to future improvements for addressing limitations, further exploration can be continued based on the current result. The temporal trends identified in this paper suggest shifts in approval rates over time, but the reasons behind these changes remain unexplored. Future work could examine specific policy changes or regulatory updates to understand their impact on approval trends. Additionally, external events such as economic downturns, housing crises, or infrastructure investments could be analyzed for their influence on application volumes and decisions. A difference-in-differences approach could also help isolate the effects of significant policy interventions, providing a clearer picture of causal relationships.

Given the importance of geographic variability in approval rates, future research could adopt spatial analysis techniques. Spatial regression models could account for spatial autocorrelation and clustering effects in planning districts, while network analysis could study relationships between districts, such as spillover effects of development activity. Additionally, analyzing proximity to key infrastructure, such as transit hubs or commercial centers, could provide insights into spatial patterns of decision-making.

@@

In summary, future research should prioritize richer data integration, explore more sophisticated modeling techniques, and broaden the scope of analysis to other regions and temporal contexts. These steps would not only enhance academic understanding of urban planning decisions but also provide actionable insights for policymakers, developers, and urban planners working toward effective and equitable processes.

@@

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This comparison highlights how the data influenced parameter estimates. Large shifts from the prior to the posterior, such as seen for certain planning districts and years, suggest strong data-driven updates. Conversely, parameters where the posterior closely resembles the prior indicate minimal influence from the observed data. This evaluation helps assess the informativeness of the priors and the robustness of the parameter estimates, ensuring the model is both data- and prior-consistent.

@@

@fig-stanareyouokay-1 is a trace plot. It shows he sampled values for each chain of the Markov Chain Monte Carlo (MCMC) simulation over iterations. The chains appear to mix well and explore the parameter space without any evident trends or stickiness, suggesting convergence. The consistent overlap of chains indicates that the posterior distribution has been effectively sampled, reducing the risk of bias due to poor chain mixing or non-convergence.

@fig-stanareyouokay-2 is a Rhat plot. It evaluates the Gelman-Rubin convergence diagnostic. All parameters show $\hat{R}$ values close to 1, which indicates that the between-chain and within-chain variances are nearly identical. This confirms that the MCMC chains have converged and are sampling from the same target distribution, validating the reliability of the posterior estimates.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 1

plot(model, "trace")

@@

Within these strata, respondents will be randomly selected to align with the distribution of these categories in the broader population. The process begins by grouping the population into strata defined by characteristics such as application type, planning district, and applicant role. Afterward, random samples are drawn independently from each subgroup, ensuring proportional representation. In cases where smaller groups are critical to the study, adjustments can be made to ensure adequate representation. Finally, the selected samples from each subgroup are combined to create the overall sample, ensuring that all significant categories are included.

@@

I appreciate your participation in this survey, which seeks to better understand the factors influencing urban planning decisions and application outcomes in Toronto. Your responses are crucial for enhancing the research and contributing to insights that could inform future urban planning processes.

@@

Thank you for your valuable contribution! As a token of our appreciation, each participant will receive $5 upon completion of the survey.

@@

Do you have any further comments or insights about the factors that might affect the application decision?

@@

Do you have any further comments or insights about the application process?

@@

We sincerely appreciate your time and thoughtful responses. Your participation is essential in helping us gain a deeper understanding of the factors influencing urban planning decisions in Toronto. Your insights will contribute to research aimed at improving zoning application processes and understanding decision dynamics.

@@

As a token of our gratitude, each participant will receive a $5 reward shortly after completing the survey. Thank you again for your valuable contribution!

@@

Question phrasing was tailored to align with terminology familiar to the target respondents, minimizing ambiguity. Efforts were also made to limit the survey's length to encourage participation while ensuring sufficient data for meaningful analysis. A balance was struck between closed-ended questions, which facilitate quantitative analysis, and open-ended questions, which capture nuanced qualitative insights. Additionally, the survey design accounted for ethical considerations, such as maintaining confidentiality and emphasizing voluntary participation.

@@

The survey design has several strengths. Stratified random sampling ensures that the results are representative of the population, reducing sampling bias. By aligning strata with key demographic and process variables, the survey facilitates detailed subgroup analysis, enhancing the depth of insights. The inclusion of both quantitative and qualitative questions allows for an overall understanding of the factors influencing urban planning decisions. Closed-ended questions enable statistical analysis, while open-ended responses provide richer context and identify factors not captured in the observational dataset.

The survey also addresses gaps in the cleaned dataset, such as perceptions of community feedback, challenges during the application process, and external influences on decisions. These insights complement the quantitative data, offering a more holistic view of the decision-making process.

@@

The survey's limited length, while improving completion rates, may constrain the depth of responses, especially for complex topics. Furthermore, the survey results are context-specific, reflecting Toronto’s unique regulatory framework and urban planning dynamics, which may limit the generalizability of findings to other cities or contexts. Despite these challenges, the survey is a valuable tool for supplementing the dataset and addressing research objectives.

>>>>

---
title: "Understanding Factors Contributing to Application Approval in Urban Planning Decisions (2016-2024)"
subtitle: "Analysis using data from Toronto’s Committee of Adjustment applications using Bayesian Logistic Regression"
author: Angel Xu
thanks: "Code and data are available at: [https://github.com/Anjojoo/Committee-of-Adjustment-Applications](https://github.com/Anjojoo/Committee-of-Adjustment-Applications)."
date: today
date-format: long
abstract: "This paper analyzes urban planning application approvals submitted to Toronto's Committee of Adjustment from 2016 to 2024, focusing on key predictors such as application type, submission year, and planning district using Bayesian logistic regression. The analysis shows that consents are more likely to be approved than minor variances, with approval rates generally increasing over time, particularly from 2017 onward, though fluctuations are observed in certain years. Geographic disparities were identified, with North York showing the highest likelihood of approval, while Scarborough faced stricter reviews. These findings enhance understanding of urban planning decision-making processes, offering perceptions into temporal trends, geographic variability, and areas for equitable policy adjustments."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

@@

This study aims to estimate the factors influencing the approval decisions for urban planning applications in Toronto. By analyzing variables such as the type of application (minor variance or consent), the year of submission, and the planning district, this study seeks to understand how these elements shape the decision-making process. The primary objective is to determine the causal effects of these factors on approval outcomes, thereby offering perceptions into the dynamics of urban planning and regulatory processes.

@@

The Committee of Adjustment plays a key role in Toronto’s urban planning ecosystem. It ensures that deviations from zoning bylaws align with the city’s broader development objectives, maintaining a balance between flexibility for property owners and adherence to long-term planning goals. The decisions made by the committee are influenced by various factors, including the type of application, geographic considerations (e.g. planning districts), and the timing of submission. Understanding these decisions provides awareness of urban planning dynamics and the regulatory challenges faced by a growing city like Toronto.

@@

The dataset thus provides a structured lens through which to examine urban planning decisions. By distilling real-world phenomena—such as zoning adjustments and land use approvals—into discrete variables, the dataset enables systematic analysis of trends and relationships within Toronto’s regulatory framework. This structured approach ensures that the dataset is both practical for analysis and representative of the detailed processes underlying urban development.

@@

#### Plot the distribution of `decision` ####
ggplot(analysis_data, aes(x = factor(decision))) +
  geom_bar(fill = "lightblue", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -1.0) +
  labs(
    x = "Decision (0 = Refused, 1 = Approved)",
    y = "Count"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

@@

#### Plot the distribution of `application_type` ####
ggplot(analysis_data, aes(x = factor(application_type))) +
  geom_bar(fill = "lightgreen", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -1.0) +
  labs(
    x = "Application Type",
    y = "Count"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

@@

The year variable represents the year of submission for each application in the dataset. It provides awareness into temporal trends in applications submitted to the Committee of Adjustment. This variable is instrumental in identifying changes in the volume of applications over time, which may reflect policy shifts, urban development trends, or external factors like economic conditions.

@@

#### Plot the distribution of `year` ####
ggplot(analysis_data, aes(x = factor(year))) +
  geom_bar(fill = "skyblue", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.8) + # Move labels higher
  labs(
    x = "Year",
    y = "Count"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) + # Add extra space above bars
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12)
  )
```

@@

The year variable is likely related to the decision outcome and application type. As approval rates may have varied over time due to changing city planning policies or committee priorities. Besides, temporal trends may highlight shifts in the proportion of Minor Variances (MV) versus Consents (CO), reflecting evolving development demands.

@@

The planning district variable identifies the geographic district within Toronto where an application was submitted. The city is divided into four planning districts, each with its own development characteristics and zoning regulations. These districts are pivotal for understanding spatial trends in application submissions and outcomes, as local policies and priorities can influence decision-making processes.

@@

#### Plot the distribution of `planning_district` ####
ggplot(analysis_data, aes(x = factor(planning_district))) +
  geom_bar(fill = "coral", color = "black") +
  geom_text(stat = "count", aes(label = ..count..), vjust = -1.0) + # Move labels higher
  labs(
    x = "Planning District",
    y = "Count"
  ) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) + # Add extra space above bars
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    axis.text = element_text(size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis text
  )
```

@@

After observing the data, the posterior distributions of $\beta$ are obtained by combining these priors with the likelihood of the observed outcomes. The resulting posterior distributions provide updated beliefs about the coefficients, incorporating both prior knowledge and evidence from the data. This approach allows for robust inference, offering a clear quantification of parameter uncertainty and supporting more detailed interpretations of the predictors' effects.

@@

The Bayesian logistic regression model is an appropriate choice for analyzing how application type, submission date, and planning district influence the likelihood of application approval. This choice is driven by the binary nature of the response variable, the model's ability to incorporate prior knowledge, and its probabilistic framework, which quantifies uncertainty more effectively than frequentist alternatives. Logistic regression naturally constrains predicted probabilities to the [0,1] interval, ensuring interpretability as likelihoods of approval, while the Bayesian approach enhances flexibility and provides richer revelations through posterior distributions.

@@

This Bayesian model relies on several assumptions that ensure the validity of its predictions and the transparency of its coefficients. While these assumptions are generally less restrictive than those for simpler models like linear regression, they remain important to the robustness of the results. The key assumptions are listed below:

@@

The result highlights how the predictors influence the likelihood of application approval while controlling for other variables. The coefficients, presented as log-odds, provide observations into both the direction and magnitude of each predictor’s impact on the approval decision.

@@

The year variables display significant temporal variations in approval likelihood, using 2016 as the reference year. These coefficients highlight how approval dynamics have shifted over time. For instance, in 2017, the coefficient of 0.74 indicates a 109% increase in the odds of approval compared to 2016 ($e^{0.74} \approx 2.09$), which suggests a substantial policy or procedural shift. In 2018, the strongest positive effect is observed, with a coefficient of 0.96, reflecting a 161% increase in odds relative to 2016 ($e^{0.96} \approx 2.61$). However, in 2019, the coefficient drops to 0.22, representing a smaller increase of 25% in odds ($e^{0.22} \approx 1.25$), suggesting that this period may have faced stricter reviews or procedural adjustments. By 2020, the coefficient rebounds to 0.56, translating to a 75% increase in odds ($e^{0.56} \approx 1.75$).

@@

In summary, the model presents key patterns in how application type and submission year influence the likelihood of approval. Minor variance applications face slightly lower odds of approval compared to consents, while temporal trends indicate an overall increase in approval rates from 2016 to 2020. These findings provide a foundation for understanding the dynamics of approval decisions and highlight areas for further investigation, such as geographic variability or interactions between predictors.

@@

This paper investigates the factors influencing application approvals submitted to the City of Toronto's Committee of Adjustment, using a Bayesian logistic regression model. By analyzing predictors such as application type, submission year, and planning district, the study provides findings into the dynamics of urban planning decisions. The model captures temporal trends, spatial variability, and procedural differences, offering a structured understanding of approval outcomes. Visualizations, including coefficient plots and effect plots, complement the analysis, helping to interpret the relationships between predictors and approval likelihood.

@@

One key finding is the temporal trend in approval rates. The analysis exhibits that approval odds have increased steadily over time, peaking in 2018. This trend may reflect evolving planning policies, greater efficiency in committee processes, or a shift in zoning priorities. The data suggests that the Committee of Adjustment has become more permissive in granting approvals, which could indicate a growing alignment between applications and city planning objectives. This result highlights how administrative processes adapt over time to urban development needs.

@@

Spatial variability plays a significant role in application approvals. The findings show that North York consistently exhibits the highest approval rates, likely due to its prominence as a development hub with infrastructure supporting large-scale projects. Conversely, Scarborough has the lowest odds of approval, suggesting stricter zoning enforcement or less alignment with development objectives in this area. This geographic disparity underscores the influence of local contexts and priorities on urban planning decisions, manifesting how different districts balance growth and zoning compliance.

@@

While this paper provides findings into the factors influencing application approvals, there exists several limitations. These weaknesses cover data-related issues, methodological constraints, and broader contextual considerations.

1. **Data Limitations**
The analysis relies solely on the dataset provided by the City of Toronto's Committee of Adjustment, which, while complete in its coverage of applications, lacks certain important variables. Key contextual factors, such as community objections, project complexity, or the presence of political or economic pressures, are not captured. For instance, the dataset does not provide information about why certain applications are refused or approved beyond the recorded decision, limiting the ability to analyze qualitative or situational influences.

@@

3. **Simplified Representation of Approval Dynamics**
The approval process involves a combination of bureaucratic procedures, stakeholder negotiations, and technical considerations that are not fully captured in this analysis. For example, public hearings, lobbying efforts, and input from local communities likely play a significant role in shaping decisions but are absent from the dataset. This simplification means that the model captures only part of the story, potentially overlooking sophisticated dynamics in urban planning and governance.

@@

One important next step is to expand the dataset by incorporating additional variables that capture contextual and qualitative factors influencing decisions. Data on community feedback and objections during public hearings is a good choice since it could show how local resistance or support impacts approval likelihood. Information on project complexity, such as the scale, type, and potential impact of proposals, could indicate patterns in decision-making. Besides, access to committee meeting notes or further investigation in meeting minutes would help clarify the reasoning behind specific approvals or refusals. Furthermore, integrating economic and political context, such as local economic indicators or changes in political leadership, would offer a broader understanding of approval trends.

The current model assumes additive relationships between predictors and approval likelihood, which may oversimplify real-world dynamics. Future research could explore nonlinear relationships, such as whether certain predictors have diminishing or exponential effects on approval rates. Interaction effects between predictors, such as the interplay between planning districts and application types, could also provide more subtle observation. For instance, minor variances may be treated differently in densely populated urban areas compared to suburban districts.

@@

Despite to future improvements for addressing limitations, further examination can be continued based on the current result. The temporal trends identified in this paper suggest shifts in approval rates over time, but the reasons behind these changes remain unexplored. Future work could examine specific policy changes or regulatory updates to understand their impact on approval trends. Additionally, external events such as economic downturns, housing crises, or infrastructure investments could be analyzed for their influence on application volumes and decisions. A difference-in-differences approach could also help isolate the effects of significant policy interventions, providing a clearer picture of causal relationships.

Given the importance of geographic variability in approval rates, future research could adopt spatial analysis techniques. Spatial regression models could account for spatial autocorrelation and clustering effects in planning districts, while network analysis could study relationships between districts, such as spillover effects of development activity. Additionally, analyzing proximity to key infrastructure, such as transit hubs or commercial centers, could provide perceptions into spatial patterns of decision-making.

@@

In summary, future research should prioritize richer data integration, explore more sophisticated modeling techniques, and broaden the scope of analysis to other regions and temporal contexts. These steps would not only enhance academic understanding of urban planning decisions but also provide actionable perceptions for policymakers, developers, and urban planners working toward effective and equitable processes.

@@

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This comparison highlights how the data influenced parameter estimates. Large shifts from the prior to the posterior, such as seen for certain planning districts and years, suggest strong empirical updates. Conversely, parameters where the posterior closely resembles the prior indicate minimal influence from the observed data. This evaluation helps assess the informativeness of the priors and the robustness of the parameter estimates, ensuring the model is both data- and prior-consistent.

@@

@fig-trace presents trace plots. It shows the sampled values for each chain of the Markov Chain Monte Carlo (MCMC) simulation over iterations. The chains appear to mix well and explore the parameter space without any evident trends or stickiness, suggesting convergence. The consistent overlap of chains indicates that the posterior distribution has been effectively sampled, reducing the risk of bias due to poor chain mixing or non-convergence.

@fig-rhat is a Rhat plot. It evaluates the Gelman-Rubin convergence diagnostic. All parameters show $\hat{R}$ values close to 1, which indicates that the between-chain and within-chain variances are nearly identical. This confirms that the MCMC chains have converged and are sampling from the same target distribution, validating the reliability of the posterior estimates.

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-trace
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot of Intercept", "Trace plot of Application type MV", "Trace plot of year 2017", "Trace plot of year 2018", "Trace plot of year 2019", "Trace plot of year 2020", "Trace plot of year 2021", "Trace plot of year 2022", "Trace plot of year 2023", "Trace plot of year 2024", "Trace plot of planning district NorthYork", "Trace plot of planning district Scarborough", "Trace plot of planning district EastYork"]
#| layout-ncol: 3

plot(model, "trace", "(Intercept)")
plot(model, "trace", "application_typeMV")
plot(model, "trace", "year2017")
plot(model, "trace", "year2018")
plot(model, "trace", "year2019")
plot(model, "trace", "year2020")
plot(model, "trace", "year2021")
plot(model, "trace", "year2022")
plot(model, "trace", "year2023")
plot(model, "trace", "year2024")
plot(model, "trace", "planning_districtNorth York")
plot(model, "trace", "planning_districtScarborough")
plot(model, "trace", "planning_districtToronto East York")

```

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-rhat
#| fig-cap: "Rhat plot"
#| layout-ncol: 1

@@

Within these strata, respondents will be randomly selected to align with the distribution of these categories in the broader population. The process begins by grouping the population into strata defined by characteristics such as application type, planning district, and applicant role. Afterward, random samples are drawn independently from each subgroup, ensuring proportional representation. In cases where smaller groups are important to the study, adjustments can be made to ensure adequate representation. Finally, the selected samples from each subgroup are combined to create the overall sample, ensuring that all significant categories are included.

@@

I appreciate your participation in this survey, which seeks to better understand the factors influencing urban planning decisions and application outcomes in Toronto. Your responses are pivotal for enhancing the research and contributing to perceptions that could inform future urban planning processes.

@@

Thank you for your significant contribution! As a token of our appreciation, each participant will receive $5 upon completion of the survey.

@@

Do you have any further comments or awareness about the factors that might affect the application decision?

@@

Do you have any further comments or awareness about the application process?

@@

We sincerely appreciate your time and thoughtful responses. Your participation is essential in helping us gain a deeper understanding of the factors influencing urban planning decisions in Toronto. Your awareness will contribute to research aimed at improving zoning application processes and understanding decision dynamics.

@@

As a token of our gratitude, each participant will receive a $5 reward shortly after completing the survey. Thank you again for your significant contribution!

@@

Question phrasing was tailored to align with terminology familiar to the target respondents, minimizing ambiguity. Efforts were also made to limit the survey's length to encourage participation while ensuring sufficient data for meaningful analysis. A balance was struck between closed-ended questions, which facilitate quantitative analysis, and open-ended questions, which capture intricate qualitative findings. Additionally, the survey design accounted for ethical considerations, such as maintaining confidentiality and emphasizing voluntary participation.

@@

The survey design has several strengths. Stratified random sampling ensures that the results are representative of the population, reducing sampling bias. By aligning strata with key demographic and process variables, the survey facilitates detailed subgroup analysis, enhancing the depth of discovery. The inclusion of both quantitative and qualitative questions allows for an overall understanding of the factors influencing urban planning decisions. Closed-ended questions enable statistical analysis, while open-ended responses provide richer context and identify factors not captured in the observational dataset.

The survey also addresses gaps in the cleaned dataset, such as perceptions of community feedback, challenges during the application process, and external influences on decisions. They complement the quantitative data, offering a more holistic view of the decision-making process.

@@

The survey's limited length, while improving completion rates, may constrain the depth of responses, especially for complex topics. Furthermore, the survey results are context-specific, reflecting Toronto’s unique regulatory framework and urban planning dynamics, which may limit the generalizability of findings to other cities or contexts. Despite these challenges, the survey is a useful tool for supplementing the dataset and addressing research objectives.
